If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-13 GroupPolarizationStatmod/paper/intro-latest.tex
revised streamlined intro

\begin{itemize}
\tightlist
\item
  We demonstrate here that there is a pervasive problem observed over
  decades of group polarization research where faulty statistical
  methods and incomplete reporting that should force the retraction of
  these studies from the canon. Effectively, then, empirical support for
  the theoretical concept \emph{group polarization} stops being so: the
  epistemic integrity of \emph{group polarization} is undermined with
  its empirical foundation losing mass.
\end{itemize}

We complement this critical analysis with a prescriptive analysis to
demonstrate that more appropriate statistical models do indeed lower the
false discovery rate by committing fewer false positive detections of
group polarization under a null model with shrinking opinion variance.
Recently, statistical best-practice is to use Bayesian models of
statistics, where social dynamics and opinion-reporting behaviors are
represented as causally-dependent probability distributions, where a
sample taken from one distribution (e.g., the opinion distribution of a
group in a group polarization experiment) acts as an input parameter to
another distribution (e.g., that represents the
individually-heterogeneous distortion of opinions when reported on an
ordinal scale). Our models follow current best practices in
psychometrics, social science, and Bayesian inference (Liddell and
Kruschke 2018; Kruschke 2015; McElreath 2020; Deffner et al. 2024), but
consolidate them in a new, useful way that enables the retroactive
discovery of null model parameters that generate spurious group
polarization, forcing published evidence for group polarization to be
classified as non-identifiable, and removed from evidence that can be
said to support group polarization.

Our methods can be turned on any data-driven social science analysis of
data where participants report a latent, subjective psychological
variable (mood, physiological state, etc.), and where various theories
supposedly battle it out to best represent the phenomenon of interest
via NHST performed on ordinal observations with weak, informal
theoretical models to motivate the studies. This may just be the first
of many bricks to fall since \emph{all 68} of the articles that
mentioned the word ``Likert'' and analyzed ordinal data used metric
models to detect latent psychological variable change across three
influential psychology journals, \emph{Psychological Science}, the
\emph{Journal of Experimental Psychology: General}, and the
\emph{Journal of Personality and Social Psychology} (Liddell and
Kruschke 2018).

This work is motivated by a core conviction that better thinking leads
to better science. Theoretical clarity, careful measurement, and
principled statistical modeling are not academic luxuries, but essential
tools for empirical science. Theoretical and statistical confusions in
the group polarization literature have lain dormant, undermining
measurements of opinion change in group polarization unbeknownst to
generations of researchers (Brown 1986).

>>>>

% \begin{itemize}
% \tightlist
% \item
We developed a generative model of opinion change to identify, if possible,
null opinion change models that could spuriously appear like group polarization
when naively measured using a Likert scale without accounting for the measurement procedure. 
We further applied our method to estimate the family-wise error rate of
detections of group polarization in a corpus of ten influential papers from 
across several decades and from various theoretical traditions. We turn this criticism of past
blunders constructive by providing a procedure to calculate what Cohen's $d$ should be
used to guarantee a family-wise error rate (i.e., Type I error rate) of 5\% or less.

We calculate the severity of the problem in two steps. In the first step we 
provide proof by example that there exists a null opinion change model in terms of \emph{latent, continuous}
psychological opinions that looks like group polarization when measured via participant behavior, i.e.,
by their response on an \emph{ordinal survey scale}. If such a null model is found, this means that we cannot
rule out a null effect, given the published data. This means, therefore, that the published detection is
just as likely false as it is true, so we say it is a \emph{plausibly false detection of group polarization},
or a \emph{plausibly spurious detection} for short.

In the second step of the analysis of whether, we simulate observations
of group polarization experiments where participant opinions are drawn from the 
null model distributions identified in the first step,
then these are measured on a Likert scale for simulated group polarization survey
responses on an ordinal scale. We then calculate the
family-wise error rate (i.e., Type I error rate) and false detection rate using
various effect sizes to determine significance of the effect in terms of Cohen's $d$.
The family-wise error rate is defined as the probability that a known null 
effect is detected as significant. We use this to then estimate the *false detection rate*, which is the
probability that a detection of an effect is a false one, which depends on the relative prevalence of
true and false positive detections in the literature.
Finally, we put our model to constructive use to show what Cohen's $d$ must be set to
to limit the family-wise error rate to 5\%, denoted $d*$. This will not help rescue any of
the published results here found to be plausibly spurious. But, it provides a measure of the weakness of
the published experimental designs independent of their shared fatal failure to 
account for the measurement procedure: greater $d*$ indicates weaker designs.

We find that 95\% (54 of 57) of group polarization detections are plausibly spurious.
In other words, empirical support for group polarization has been seriously undermined
just by removing these results from the count of group polarization replications.
There is no reason to expect other papers to be more reliable. (Summarize other
two findings, 1. FWER/FDR and 2. $d*$ for FWER < 5\%)

Our methods can be turned on any data-driven social science analysis of
data where participants report a latent, subjective psychological
variable (mood, physiological state, etc.), and where various theories
supposedly battle it out to best represent the phenomenon of interest
via null-hypothesis significance testing using ordinal data as continuous. 
This may just be the first
of many bricks to fall since \emph{all 68} of the articles that
mentioned the word ``Likert'' and analyzed ordinal data used metric
models to detect latent psychological variable change across three
influential psychology journals, \emph{Psychological Science}, the
\emph{Journal of Experimental Psychology: General}, and the
\emph{Journal of Personality and Social Psychology} (Liddell and
Kruschke 2018). This further confirms that replication is far from the only
foundational crisis that behavioral science faces\cite{Yarkoni2022,Turner2022}.

Replication is not enough:
strong, mechanistic models must be developed to be able to predict Type I and II
error rates *ahead of time*, as we demonstrate here. We only demonstrate how to
calculate expected Type I error rates for a measurement design, i.e., the rate of false 
positives. Low statistical power exacerbates false detection rates. So, to be confident in replication,
a pre-registered measurement design should use this style of simulation to 
estimate the statistical power. Tangled, weak, informal
verbal models of complex psychological and social phenomena like those that motivated
empirical studies of group polarization tend to be vulnerable
to stastical weaknesses. It is difficult to incorporate prior knowledge such
as experiment structure without a sufficiently rigorous and appropriate 
formal system \cite{Smaldino2017,Smaldino2020,Flake2020,Deffner2024}.

% This work is motivated by a core conviction that better thinking leads
% to better science. Theoretical clarity, careful measurement, and
% principled statistical modeling are not academic luxuries, but essential
% tools for empirical science. Theoretical and statistical confusions in
% the group polarization literature have lain dormant, undermining
% measurements of opinion change in group polarization unbeknownst to
% generations of researchers (Brown 1986).
